---
title: "hw2"
author: "Min Jegal, Alba Arribas"
date: "2023-12-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---
title: "homework2"
author: "Min Jegal, Alba Arribas"
date: "2023-12-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Classification and regression

For this assignment we are using a wine database https://archive.ics.uci.edu/dataset/186/wine+quality. This database is divided in two, both of them with the same 12 variables. We have the red and white database which later on we will combine. 
this database consist on the following values:
- fixed acidity : provides a feeling of freshness and liveliness on the palate
- volatile acidity : perceived as the odour of vinegar
- citric acid : acid supplement that winemakers use to boost the acidity of their wine especially grapes grown in warmer climates
- residual sugar
- chlorides
- free sulfur dioxide
- total sulfur dioxide
- density
- pH
- sulphates
- alcohol
- quality (score between 0 and 10)

So our main goal here will be to determine which of this factors affect the wine the most and how it differs from one wine type to another.

```{r cars}
setwd("A:/AI_STUDY/Practice/2023-2024 MADRID/Statistical Learning/HW2/wine+quality")
white_wine = read.csv(file = "winequality-white.csv", header=T, sep=";", dec=".")
red_wine = read.csv(file = "winequality-red.csv", header=T, sep=";", dec=".")
head(white_wine)
```
Here, we combinen both databases adding new column called type with a categorical variable indicating if the wine is red or white.


```{r pressure, echo=FALSE}
white_wine$type <- "white"
red_wine$type <- "red"

# Combine the two dataframes
wine.df <- rbind(white_wine, red_wine)
head(wine.df)
```

```{r pressure, echo=FALSE}
summary(wine.df)
dim(wine.df)
str(wine.df)
```
Now we check if there are any missing values.
```{r pressure, echo=FALSE}
barplot(colMeans(is.na(wine.df)), las=2)
barplot(c(any(is.na(wine.df))), las=2)
```
After that, we check for any outliers that don't really make sense.

```{r pressure, echo=FALSE}

library(outliers)
boxplot(subset(wine.df, select = -c(13)))$out
boxplot(subset(wine.df, select = -c(6,7,13)))$out
```

We can see that there are not negative values and the outliers we have might indicate that a wine is specially good or specially bad therefore getting rid of them will result in a loss of information. Therefore, to handle this outliers we could just standardize the code so it minimizes the distortion.


```{r pressure, echo=FALSE}
library(corrplot)
correlation_matrix <- cor(wine.df[, -which(names(wine.df) == "type")])
corrplot(correlation_matrix, method = "color", tl.cex = 0.7, tl.col = "black", addrect = 2)
```

From this matrix we can conclude that almost all variables are correlated in way but there are some that stand out such as the positive correlation between the total.sulfur.dioxide and the free.sulfur.dioxide which could make sense and in the other hand, the alcohol and the sensitive are negatively correlated.


```{r pressure, echo=FALSE}
library(ggplot2)
library(lmtest)

ggplot(wine.df, aes(x = alcohol, y = quality, color = type)) +
  geom_point() +
  labs(x = "Alcohol", y = "Quality")
```
```{r pressure, echo=FALSE}

# Fit regression model
reg_model <- lm(quality ~ alcohol + pH + sulphates + type, data = wine.df)

# Check regression assumptions
# Residuals vs Fitted plot
residuals_vs_fitted <- ggplot(reg_model, aes(.fitted, .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residuals vs Fitted Values",
       x = "Fitted Values",
       y = "Residuals")

# Normal Q-Q plot
qq_plot <- ggplot(reg_model, aes(sample = .stdresid)) +
  stat_qq() +
  stat_qq_line() +
  labs(title = "Normal Q-Q Plot of Residuals")

print(residuals_vs_fitted)
print(qq_plot)

# Run diagnostic tests
# Breusch-Pagan test for heteroscedasticity
bp_test <- bptest(reg_model)
print(bp_test)

# Shapiro-Wilk test for normality of residuals
shapiro_test <- shapiro.test(reg_model$residuals)
print(shapiro_test)


```


```{r pressure, echo=FALSE}

# Assuming 'df' is your data frame
# Install and load the ggplot2 library if not already installed
# install.packages("ggplot2")
library(ggplot2)

# Create a histogram using ggplot2
ggplot(wine.df, aes(x = residual.sugar)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Residual Sugar",
       x = "Residual Sugar",
       y = "Frequency")
```