---
title: "hw2"
author: "Min Jegal, Alba Arribas"
date: "2023-12-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---
title: "homework2"
author: "Min Jegal, Alba Arribas"
date: "2023-12-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Classification and regression

For this assignment we are using a wine database https://archive.ics.uci.edu/dataset/186/wine+quality. This database is divided in two, both of them with the same 12 variables. We have the red and white database which later on we will combine. 
this database consist on the following values:

**fixed acidity** : most acids involved with wine or fixed or nonvolatile (do not evaporate readily)

**volatile acidity** : the amount of acetic acid in wine, which at too high of levels can lead to an unpleasant, vinegar taste

**citric acid** : found in small quantities, citric acid can add 'freshness' and flavor to wines

**residual sugar** : the amount of sugar remaining after fermentation stops, it's rare to find wines with less than 1 gram/liter and

**chlorides** : the amount of salt in the wine

**free sulfur dioxide** : the free form of SO2 exists in equilibrium between molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents

**total sulfur dioxide** : amount of free and bound forms of S02; in low concentrations, SO2 is mostly undetectable in wine, but at free SO2

**density** : the density of water is close to that of water depending on the percent alcohol and sugar content

**ph** : describes how acidic or basic a wine is on a scale from 0 (very acidic) to 14 (very basic); most wines are between 3-4 on the

**sulphates** : a wine additive which can contribute to sulfur dioxide gas (S02) levels, wich acts as an antimicrobial and

**alcohol sulfur dioxide** : the percent alcohol content of the wine

**quality** : output variable (based on sensory data, score between 0 and 10)

So our main goal here will be to determine which of this factors affect the wine the most and how it differs from one wine type to another.

```{r cars}
setwd("A:/AI_STUDY/Practice/2023-2024 MADRID/Statistical Learning/HW2/Statistical-Learning_Wine/wine+quality")
white_wine = read.csv(file = "winequality-white.csv", header=T, sep=";", dec=".")
red_wine = read.csv(file = "winequality-red.csv", header=T, sep=";", dec=".")
head(white_wine)
```
Here, we combine both databases adding new column called type with a categorical variable indicating if the wine is red or white.


```{r pressure, echo=FALSE}
white_wine$type <- "white"
red_wine$type <- "red"

# Combine the two dataframes
wine.df <- rbind(white_wine, red_wine)
head(wine.df)
```

```{r pressure, echo=FALSE}
summary(wine.df)
dim(wine.df)
str(wine.df)
```
Now we check if there are any missing values.
```{r pressure, echo=FALSE}
barplot(colMeans(is.na(wine.df)), las=2)
barplot(c(any(is.na(wine.df))), las=2)
```
Seems like there is no NAs.
After that, we check for any outliers. Thinking of the data, the outliers will detect few excellent or poor wines.

```{r pressure, echo=FALSE}

library(outliers)
boxplot(subset(wine.df, select = -c(13)))$out
boxplot(subset(wine.df, select = -c(6,7,13)))$out
```

We can see that there are not negative values and the outliers we have might indicate that a wine is specially good or specially bad therefore getting rid of them will result in a loss of information. Therefore, to handle this outliers we could just standardize the code so it minimizes the distortion.


```{r pressure, echo=FALSE}
library(corrplot)
correlation_matrix <- cor(wine.df[, -which(names(wine.df) == "type")])
corrplot(correlation_matrix, method = "color", tl.cex = 0.7, tl.col = "black", addrect = 2)
```

From this matrix we can say that almost all variables are correlated in way but there are some that stand out such as the positive correlation between the total.sulfur.dioxide and the free.sulfur.dioxide which could make sense and in the other hand, the alcohol and the sensitive are negatively correlated.


## Classification

1. Type Prediction : classify wines based on their type
2. Quality Prediction : to predict the quality of the wine based on its physicochemical properties. 









```{r pressure, echo=FALSE}
library(ggplot2)
library(lmtest)

ggplot(wine.df, aes(x = alcohol, y = quality, color = type)) +
  geom_point() +
  labs(x = "Alcohol", y = "Quality")

```
According to the chart, it seems like the mid ragned quailty wines (5 - 7) shows a big variablity on the strongness of Alcohol, however, the low quality wines have lower Alcohol than the high quality wines.
```{r}
library(dplyr)

columns_to_plot <- c("fixed.acidity", "volatile.acidity", "citric.acid", "residual.sugar", 
                     "chlorides", "free.sulfur.dioxide", "total.sulfur.dioxide", 
                     "density", "pH", "sulphates", "alcohol")

create_plot <- function(column_name) {
  grouped_data <- wine.df %>%
    group_by(quality) %>%
    summarise(mean_value = mean(get(column_name), na.rm = TRUE)) %>%
    arrange(quality)

  ggplot(grouped_data, aes(x = quality, y = mean_value, fill = mean_value)) +
    geom_text(aes(label = round(mean_value, 2)), vjust = -0.3, size = 3) +
    geom_col() +
    scale_fill_gradient(low = "white", high = "red") +
    theme_minimal() +
    labs(title = paste("Mean", column_name, "by Quality"), x = "Quality", y = paste("Mean", column_name))
}

plots <- lapply(columns_to_plot, create_plot)
print(plots)
```
We can say that the increase in fixed acidity positively affects the given vote.but it is difficult to make a full conclusion at this stage.
we can make a more precise sentence according to the comment we made above. The decrease in volatile acidity affects the votes positively.
we see that the increase in citric acid positively affected the votes
It seems difficult to make an inference about residual sugar from here.
The decrease in chlorides positively affects the votes cast.
It doesn't seem easy to comment on free sulfur dioxide either. The 13, 14 levels seem to have been voted well.
It doesn't make sense to comment on total sulfur dioxide. It changed according to the votes. There is no order.
The intensity is almost the same in all votes, but there is only a slight decrease. Density drop seems to have a positive effect on the votes.
We can say that the decrease in pH affects the votes positively.
It is clear from here that the increase in sulphates has a positive effect.
In general, it can be said that the increase in alcohol ratio affects the votes well.

## Distribution charts

```{r}
dist_plot <- function(column_name) {
  mean_value <- mean(data[[column_name]], na.rm = TRUE)
  sd_value <- sd(data[[column_name]], na.rm = TRUE)

  plot <- ggplot(data, aes_string(x = column_name)) +
    geom_histogram(aes(y = ..density..), binwidth = 1, fill = "#943126", color = "black") +
    geom_density(alpha = .2, fill = "#DAA520") +
    stat_function(fun = dnorm, args = list(mean = mean_value, sd = sd_value), 
                  color = "#A04000", linetype = "dashed") + # Adding normal distribution curve
    labs(title = paste("Distribution of", column_name), x = column_name, y = "Density") +
    theme_minimal()

  return(plot)
}


dist_plots <- lapply(columns_to_plot, dist_plot)
print(dist_plots)

```
For fixed acidicty we see that there is a skewness to the right. We need to fix this.

## Skewness Correction


```{r}
# Skewness Correction

```







```{r pressure, echo=FALSE}

# Fit regression model
reg_model <- lm(quality ~ alcohol + pH + sulphates + type, data = wine.df)

# Check regression assumptions
# Residuals vs Fitted plot
residuals_vs_fitted <- ggplot(reg_model, aes(.fitted, .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residuals vs Fitted Values",
       x = "Fitted Values",
       y = "Residuals")

# Normal Q-Q plot
qq_plot <- ggplot(reg_model, aes(sample = .stdresid)) +
  stat_qq() +
  stat_qq_line() +
  labs(title = "Normal Q-Q Plot of Residuals")

print(residuals_vs_fitted)
print(qq_plot)

# Run diagnostic tests
# Breusch-Pagan test for heteroscedasticity
bp_test <- bptest(reg_model)
print(bp_test)

# Shapiro-Wilk test for normality of residuals
shapiro_test <- shapiro.test(reg_model$residuals)
print(shapiro_test)


```


```{r pressure, echo=FALSE}

# Assuming 'df' is your data frame
# Install and load the ggplot2 library if not already installed
# install.packages("ggplot2")
library(ggplot2)

# Create a histogram using ggplot2
ggplot(wine.df, aes(x = residual.sugar)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Residual Sugar",
       x = "Residual Sugar",
       y = "Frequency")
```